---
title: Joe Biden Projected to Win American Popular Vote with 42.1% Predicted by a
  Logistic Regression Mixed Model
author: "Maha Mohamed"
date: "November 2,2020"
output:
  pdf_document: default
  html_document:
    df_print: paged
  word_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(skimr)
library(broom)
library(here)


setwd("C:/Users/mm84/OneDrive/Documents/School/Sta304/Problem Set 3")
# Loading in the cleaned survey Data
survey_data <- read_csv("C:/Users/mm84/OneDrive/Documents/School/Sta304/Problem Set 3/survey_data.csv")


# Loading in the cleaned census Data
census_data <- read_csv("C:/Users/mm84/OneDrive/Documents/School/Sta304/Problem Set 3/census_data.csv")



```


# GitHub repository

Code and data supporting this analysis is available at: https://popularvoteprediction.netlify.app/


# Model

   I am interested in predicting the popular vote outcome of the 2020 American federal election by using a logistic regression mixed model with post-stratification technique. 



## Model Specifics 

   This model utilizes frequentist inference as this leads to a more accurate result. The parameters of this model are fixed as its based  off survey data, thus justifying it's use (Hong et al,2013). This model utilizes survey data that was  conducted by the democracy fund voter study group in June 25 to July 1,2020 (Tausanovitch et al, 2020). It provides information regarding American voter's values and behaviors allowing us to predict who will the popular vote in the 2020 American election.The census data is from the 2018 census conducted by IPUMS providing demographics that help shape the post-stratification technique (Ruggles, 2020).
  
   A logistic regression mixed model  allows us to predict the proportion of voters who will vote for Donald Trump. This mixed model consists of fixed effects and random effects. Where, fixed effects refers to variables that will only vary between studies because of sampling error and random effects do not estimate one effect,but rather estimate the mean of a distribution of effects (Borenstein et al, 2009). In this case, the fixed effects are on the individual level as it includes sex and age. These two variables are particularly important as they provide general information about the voters allowing there to be some context to the voters. Dependent on the sex and age found more frequently and whether it's statistically significant in the context of voting for Trump, will provide valuable information regarding how Americans will vote in the upcoming election. Ages rather than age groups was utilized since age groups may not provide an accurate representation on the whole population. Sexes used the binary representation where females are 0 and males are 1. The random effect is on the group level and is the state where the household is located. This provides geographical context which plays an important role in how an American will vote as certain states are viewed to be more democratic than others (Hill et al,2019). As voting for Trump is a binary outcome this aligns with a logistic regression model further proving this is the right model for this analysis. 
  
The equation for this logistic regression mixed model provides 2 equations:

1) Individual level

$$log(\frac p{1-p})= \beta_{0j} +\beta_1 x_{sexj} +\beta_2 x_{agej}+ r_{ij}$$

Where, $$log \frac p{1-p} $$ represents the proportion of voters who will vote for Donald Trump. In this equation, j is the states where the households of the participants are located. Similarly, $\beta_0j$ represents the intercept of the model which is dependent on the random effects, in this case state. Moreover, $\beta_1$ represents the coefficient for sex of the pooled data.Therefore, for every one unit increase in sex, transitioning from female to male,we except a $\beta_1$ increase in the probability of voting for Trump. Additionally, $\beta_2$ represents the coefficient for age.So, for every one unit increase in age, we expect a $\beta_2$ increase in the probability of voting for Donald Trump. Lastly, $r_{ij}$ represents the error term for the individual level model.

2) Group level

$$ \beta_{0j}= r_{00} + r_{01} W_j +u_{0j} $$
Where, $\beta_0j$ is the intercept from the individual level equation. Additionally, $r_{00}$ is the intercept of the intercept and fixed effects determines the output of the non random terms in the random intercept model. Furthermore, $r_{01}$ is the intercept of the first state and $W_j$ is the state. Lastly, $u_{0j}$ is the error term for the group level model.


```{r, include=FALSE}

#trump results

model_logit<- glmer (as.factor(vote_trump)~sex+ age+(1|statefip),
                      data=survey_data,
                      family=binomial)

#biden results

model_biden<- glmer (as.factor(vote_biden)~sex+ age+(1|statefip),
                      data=survey_data,
                      family=binomial)


```

## Post-Stratification 

   Post-stratification is used in survey analysis to demonstrate how an entire population will vote by allowing equal allotment of subgroups of participants for each experimental condition (Downes et al, 2020). This is through  adjusting weights so that the weighted totals in cells are equal to the overall population(Downes et al,2020). This is a useful method that was followed in this analysis as it allows to account for misrepresentation in the data leading to a more accurate representation of the data. It can then be used to influence policies that accurately represent the population they are meant to serve. 
  
   Specifically, the proportion of people voting for Trump will be analyzed, once this is conducted to see the full picture,a similar analysis will be conducted for Biden. However, for simplicity we will be focusing on the proportion of people voting for Trump as the model above outlines. The cells that will be used in this stratification are age, gender, and state. Age is being included because it was seen as one of the greatest predictor in past US elections (Holland, 2013). Gender and  have also been seen to predict voting patterns as they can influence how individuals will vote as it can impact how they view policies based on gender differences (Chiao et al, 2008). Additionally, state was included since specific states tend to have similar voting patterns therefore looking at how states will votes (Hill et al,2019). Furthermore, education was excluded because studies have shown that it does not have a significant impact on voting activity as despite education levels increasing in America, voting levels remained the same (Snyder,2011).
  
   Here I made cells based off different ages, sex and states. Using the model described in the previous sub-section I will estimate the proportion of voters in each bin. I will then weight each proportion estimate in each bin by the respective population size of that bin and sum those values and divide that by the entire population size. This will lead to the proportion of voters that will vote for Trump and the methodology will be repeated for Biden.

## Data Clean up

   In order to use the census and survey data for this analysis, I first had to clean it. The survey data categories were changed to match the census data. Specifically, the mutate function was used to create the vote Trump and vote Biden categories to be able to find the proportion of voters voting for each candidate. Gender was renamed sex and females became 0's whereas males became 1's to match the census data. Within the census data, males and females were adapted from 1's and 2's to 0's and 1's. Also within census data, the state names were changed to the abbreviations found in the survey data. Age was only changed within the census data where ages that were less than 1 and 90 year old were removed this was because these ages were found not to add significant value to this analysis. Altogether, this allowed for the post-stratification analysis to work as the categories were matched.


```{r, include=FALSE}
#post-stratification calculation

#trump

census_data$logodds_estimate <-
  model_logit %>%
  predict(newdata = census_data)

census_data$estimate <-
  exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))


#biden

census_data$logodds_estimatebiden <-
  model_biden %>%
  predict(newdata = census_data)

census_data$estimatebiden <-
  exp(census_data$logodds_estimatebiden)/(1+exp(census_data$logodds_estimatebiden))



```

# Results

## Model 

  The logistic regression mixed model determining the proportion of voters that will vote for Trump showed that age, sex and state were the right variables for the model. As the fixed effects, age and sex, were statistically significant with really small p values. The random effects data shows the variance of the random effects to be 0.03.

## Table 1: This shows the results of the logistic regression model
```{r,echo=FALSE}
summary(model_logit)

```

## Post-Stratification 

The post-stratification results include the predictions for the proportion of voters who will vote for Trump and those who will vote for Biden.

## Table 2: Proportion of voters who will vote for Trump
```{r,echo=FALSE}
census_data %>% 
  mutate(apv_prop=estimate*n) %>% 
  summarise(trumppopularvote=sum(apv_prop)/sum(n))
```

## Table 3: Proportion of voters who will vote for Biden
```{r,echo=FALSE}
census_data %>% 
  mutate(apv_prop_predict=estimatebiden*n) %>% 
  summarise(bidenpopularvote=sum(apv_prop_predict)/sum(n))
```

# Discussion

   In American elections, the popular vote depicts who majority of Americans voted for. The president of the United States is declared based on the electoral college which works by Americans votes counting towards an elector who represents the candidate's party. Therefore, the popular vote plays an important role in which electors are picked who then cast the vote. The number of electors from each state is dependent on the population of the state with a minimum of 3 electors per state(Robertson et al, 2020). The electoral college system has been criticized for not accurately representing what Americans want(Robertson et al, 2020). Therefore, I am focusing solely on the predictions for the popular vote as this is solely based on who Americans want to be their president. 

   To determine the popular vote in the upcoming election, census and survey data were utilized. The survey data was  conducted by the democracy fund voter study group from June 25 to July 1,2020 providing information on American's values and voting behavior (Tausanovitch et al, 2020). As this is  recent, it allows recent data making the model accurate. The census data is from the 2018 census conducted by IPUMS providing the necessary demographic data to help shape the post-stratification technique (Ruggles, 2020).
  
   Through the logistic regression mixed model based off the survey data, I created the model that the post-stratification was based off of. The variables included within this model were age and sex as the fixed effects. The summary of the model in table 1 depicts that these variables are statistically significant further proving that this is the right model to determine the proportion of voters that will vote for Trump in the 2020 election.
  
  The following equation is for the individual level of this model:
  
  $$log(\frac p{1-p})= \beta_{0j} +0.54 x_{sexj} +0.02 x_{age}+0.09 $$
 
  The equation for the group level:

$$ \beta_{0j}= -1.59 + 0.04 W_j +0.18 $$
 
  Specifically , the equation for the individual level demonstrates that as sex transitions from female to male, there will be a 0.54 increase in the probability of voting for Trump. Illustrating that males are more likely to vote for Trump than females. This aligns with studies which show that despite females voting more than males, males tend to vote Republican whereas females are more likely to vote Democratic (Igielnik,2020). Moreover, this equation demonstrates that for every one unit increase in age, there will be a 0.02 increase in the probability of voting for Donald Trump. While this does show that older individuals are more likely to vote for Trump, this is not a large increase. This can be attributed to older voters being more likely to vote for the Democratic presidential candidate for the first time in 20 years (Stanton,2020). Specifically, the coronavirus pandemic has led many seniors to not vote for Trump as they have felt as if he has not provided them with the right support as they are a vulnerable population (Stanton,2020). The group level equation depicts the random effects model which is the state where the household of the participant is located. The variance allows for broader inference that can impact the fixed effects (Downes et al,2020). The variance was found to be 0.03 which demonstrates the variability from state to state in terms of the probability to vote for Donald Trump.


  Post stratification is utilized to provide an accurate representation of a population specifically using census data. In this analysis, it was used to determine that Joe Biden will win the popular vote by 42.1%. Whereas, Donald Trump will receive 37.2% of the popular vote in the upcoming 2020 election. This prediction demonstrates a different result from the 2016 presidential election because these last 4 years as Trump led a polarizing agenda. This has led to record high amount of people voting this year (Mellnik et al,2020). It is important to note that in 2016 despite losing the popular vote, Donald Trump became president. This was because polls did not provide an accurate representation of what Americans believed. Specifically, non response bias has been named as the cause (Mercer et al, 2016). This is when despite having equal opportunity, a group within the population decide not to respond to the poll(Mercer et al,2016). Particularly this has been associated with people not wanting to admit they support Donald Trump. 4 years later, this may still be the same Overall, the American election has implications on the entire world ad their foreign policies will impact every country.




## Weaknesses

  A weakness of this prediction is that humans can change their mind quite quickly. There's no way to know for certain how a person will vote because they can change their mind whenever they want to. While people's fundamental values tend to stay the same. There is still know way to certainly know if somebody is honestly sharing how they will vote. This can be because of the stigma surrounding voting a certain way. America within these last 4 years has changed a lot and therefore certain groups may feel as if they will be stigmatized for sharing that they will vote for a specific party. This therefore will not lead to an accurate result of who will win.



## Next Steps
  In this report, the census data was from 2018 which means that the population demographics may have changed in the last two years. This may not provide the necessary information that will influence the up and coming US election. However, this is the latest census information that is available for statistical analysis. Next steps, to lead to census data more representative of what America looks like today could be an online census. The census is trying to gain an accurate representation of the entire country which has a population of over 328 million people. Therefore, this will obviously take a long time to gather data.

  Thus, an online census that is occurring this year is a better way to adapt to this new online generation (United States Census Bureau,2020). As this online census has not yet been completed depending on what occurs it should be adapted and better moving forward. An online census can reach more individuals that will be more willing to complete it. However, considerations of the population that does not have access to technology must be accounted for. This may include making devices available that can allow them to participate in the survey. Ultimately, this leads to a better census that can lead to more accurate results within post stratification analysis, but also generally when used for policies based on the demographics of the population.



# References

1. Borenstein M, Hedges LV, Higgins JPT, Rothstein HR. 2009. Introduction to Meta-Analysis. John Wiley & Sons Ltd. ISBN: 978-0-470-05724-7.

2.  Chiao JY, Bowman NE, Gill H. 2008. The Political Gender Gap: Gender  Bias in Facial Inferences that Predict Voting Behaviour.Biochem Med. 3(2): 1-5

3.  Downes M, Carlin JB. 2020. Multilevel Regression and Poststratification as a Modeling Approach for Estimating Population Quantities in large population Health Studies: A Simulation Study. Biom J. 62(2):479-491

4.  Hansen ER, Tyner A. 2019.  Educational Attainment and Social Norms of Voting. Political Behaviour. 4(10): 9-13.

5.  Hill SJ, Hopkins DJ, Huber GA. 2019. Local Demographic Changes and US Presidential Voting, 2012 to 2016. PNAS. 116(50): 25023-25028.

6.  Holland JL. 2013. Age Gap? The Influence of Age on Voting Behaviour and Political Preferences in the American Electorate. Washington State University. 1(3):10-13.

7.  Hong H, Carlin BP, Shamliyan TA, Wyman JF, Ramakrishnan R, Sainfort F, Kane RL. 2013. Comparing Bayesian and Frequentist Approaches for Multiple Outcome Mixed Treatment Comparisons. Med Decis Making.  33(5): 702-14.


8.  Igielnik R. 2020. Men and Women in the U.S Continue to Differ in Voter Turnout Rate,  Party Identification. Pew Research Center. Available from 
https://www.pewresearch.org/fact-tank/2020/08/18/men-and-women-in-the-u-s-continue-to-differ-in-voter-turnout-rate-party-identification/

9.  Mellnik T, Berkowitz B, Santamarina D. 2020. 20 Million Americans Didn't vote in 2016's Closest Battlegrounds. The Washington Post. Available from https://www.washingtonpost.com/elections/2020/10/30/2016-nonvoters-2020-turnout/?arc404=true

10. Mercer A, Deane C, McGeeney K. 2016. Why 2016 Election Polls Missed Their Mark. Pew Research Center. Available from https://www.pewresearch.org/fact-tank/2016/11/09/why-2016-election-polls-missed-their-mark/ 

11. Robertson H, Kirk A, Hulley-Jones F. 2020.  Electoral College Explained: How Biden Faces an Uphill Battle in the US Election. Guardian. Available from  https://www.theguardian.com/us-news/ng-interactive/2020/oct/30/electoral-college-explained-how-biden-faces-an-uphill-battle-in-the-us-election 

12. Ruggles S, Flood S, Goeken R, Grover J, Meyer E, Pacas J, Sobek M. 2020. IPUMS USA: Version 10.0 [dataset]. Minneapolis, MN: IPUMS.
https://doi.org/10.18128/D010.V10.0

13. Sperandei S. 2014.  Understanding Logistic Regression Analysis. Plos One 24(1):12-18.

14. Stanton Z. 2020. Why Senior Citizens are Flipping on Trump. Politico. Available from https://www.politico.com/news/magazine/2020/10/08/senior-citizens-trump-biden-2020-voters-428131

15. Snyder RE. 2011. The Impact of Age, Education, Political Knowledge and Political Context on Voter Turnout. UNLV. 12(2): 3-5.

16. Tausanovitch  C, Vavreck L. 2020. Voter Study Group. UCLA Democracy Fund.

17. United States Census Bureau. 2020. 2018 Census Test. Available from https://www.census.gov/programs-surveys/decennial-census/2018-census-test.html



